RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-04-17 manuscripts/manuscript.Rmd
add References heading (for #112); update Dockerfile inline use notes
no need to install package, already done
edits to manuscript based on reviews

---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
  - name: Hilmar Lapp
    affiliation: NESCent
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and forward-compatible exchange standard
      recently proposed to better meet the expanding needs for
      machine-proof phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which aims to provide to users of
      the popular R programming language easy programmatic access to
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

      2. Wherever possible, we designed RNeXML to map NeXML document
      contents, whose arrangement is influenced by the format's
      normative XML Schema definition, to their most intuitive or
      useful representation in R. To make NeXML's powerful facility
      for recording semantically rich machine-readable metadata
      accessible to R users, we designed a functional programming
      interface to it that hides the semantic web standards leveraged
      by NeXML from R users who are unfamiliar with them.

@@

A detailed account of how the NeXML standard addresses these and
other relevant challenges can be found in @Vos_2012. In brief,
NeXML was designed with the following important properties. First,
NeXML is defined by a precise grammar that can be programmatically
**validated**. This ensures that any NeXML file can be expected to be
parsed and read without errors by any NeXML-compliant software, provided
the file passes validation. Second, NeXML is **extensible**: a user
can define representations of new, previously unanticipated information
(as we will illustrate) without violating its defining grammar. Third
and most importantly, NeXML is **rich in computable semantics**. The
formal semantics used in NeXML is machine-readable in a predictable
manner, and also enables machine reasoning about the data and metadata
(see @Parr2011), especially when metadata attributes and values are
anchored in appropriately axiomatized ontologies.

@@

```{r echo=FALSE, results='hide'}
install.packages("RNeXML", dependencies=TRUE)
library(RNeXML)
```

@@

Conceptually, a NeXML document has the following components: (1)
phylogeny topology data, (2) character or trait data in matrix form,
(3) operational taxonomic units (OTUs), and (4) metadata. To represent
the contents of a NeXML document (currently in memory), `RNeXML` defines
the `nexml` object type. This type therefore holds phylogenetic trees as
well as character or trait matrices, and all metadata, which is similar
to the phylogenetic data object types defined in the `phylobase` package
[@phylobase], but contrasts with the more widely used ones defined in the
`ape` package [@Paradis_2004], which represents trees alone.

@@

In addition to trees, the `nexml_write()` method also allows to specify
character data as another parameter. The following example uses data
from the comparative phylogenetics R package `geiger` [@Harmon_2008].

@@

@@

The RDFa standard adopted by NeXML (see @Vos_2012) allows for
arbitrary custom extensions to meet ongoing scientific needs without
requiring modifications (or "extensions") of the syntax. Specifically,
every metadata annotation consists of a metadata property and a value
of the property; the bearer of the property is determined by where the
annotation appears in the NeXML document (technically, the containing
element in the XML document model). This structure corresponds directly
to Resource Description Framework (RDF) subject-predicate-object triples,
and therefore the only requirement for custom metadata is that they be
cast as RDF triples.

The `RNeXML` interface described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies (such
as Dublin Core for "title", "creator", etc.) or ontologies. Giving
metadata properties (and where applicable, values) as vocabulary or
ontology terms rather than simple text strings is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` API shields her otherwise, in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

If the prefix used by the metadata property is not among the built-in
ones (which can be obtained using `get_namespaces()`), it has to be
provided along with its URI as the `namespaces` parameter. For example,
the following uses the SKOS vocabulary (which currently is not built-in)
to add a note to the trees in the `nexml` object:

@@

Because NeXML expresses all metadata using the RDF standard, and
stores them compliant with RDFa, they can be extracted as an RDF
graph, queried, analyzed, and mashed up with other RDF data, local or
on the web, using a wealth of off-the-shelf tools for working with RDF
(see @W3C_2014 or @Hartig_2012). Examples for these possibilities are
included in the `RNeXML` SPARQL vignette, and the package also comes with a
demonstration that can be run from R using the following command:
`demo("sparql", "RNeXML")`).

@@

```{r eval = FALSE}
doi <- nexml_publish(birds, visibility = "public", repository="figshare")
```

Figshare also supports "private" visibility, allowing to securely backup
data to a repository and to share them with collaborators prior to public release.

@@

>>>>

---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
  - name: Hilmar Lapp
    affiliation: NESCent
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and extensible exchange standard
      recently proposed to better meet the expanding needs for
      phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which provides users of
      the R programming language with easy-to-use tools for 
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

      2. Wherever possible, we designed RNeXML to map NeXML document
      contents, whose arrangement is influenced by the format's
      XML Schema definition, to their most intuitive or
      useful representation in R. To make NeXML's powerful facility
      for recording semantically rich machine-readable metadata
      accessible to R users, we designed a functional programming
      interface to it that hides the semantic web standards leveraged
      by NeXML from R users who are unfamiliar with them.

@@

A detailed account of how the NeXML standard addresses these and
other relevant challenges can be found in @Vos_2012. In brief,
NeXML was designed with the following important properties. First,
NeXML is defined by a precise grammar that can be programmatically
**validated**: meaning that we can verify that a file has followed this
precise grammar and thus be sure that it can be read (parsed) 
without errors by any software that uses the NeXML grammar (e.g. RNeXML). 
Second, NeXML is **extensible**: a user
can define representations of new, previously unanticipated information
(as we will illustrate) without violating its defining grammar. Third
and most importantly, NeXML is rich in **computable semantics**: patterns
that permit a machine to reason about concepts. Note that
these semantics are based on logical deduction such as: "all frogs are vertebrates,
so a query for vertebrates should return all species that are listed as frogs"), 
rather than heuristic machine learning algorithms.  Such inference requires a
computer to understand the 'semantic' meaning of a word like 'frog', which
in must be specified in a kind of dictionary known in informatics research
as an vocabulary or ontology (see @Parr2011 for a more detailed introduction
in the context of evolution and biodiversity research).

@@

```{r echo=FALSE, results='hide', eval=FALSE}
install.packages("RNeXML", dependencies=TRUE)
```

```{r echo=FALSE, results='hide'}
library(RNeXML)
```

@@

Conceptually, a NeXML document has the following components: (1)
phylogeny topology and branch length data, (2) character or trait data in matrix form,
(3) operational taxonomic units (OTUs), and (4) metadata. To represent
the contents of a NeXML document (currently in memory), `RNeXML` defines
the `nexml` object type. This type therefore holds phylogenetic trees as
well as character or trait matrices, and all metadata, which is similar
to the phylogenetic data object types defined in the `phylobase` package
[@phylobase], but contrasts with the more widely used ones defined in the
`ape` package [@Paradis_2004], which represents trees alone.

@@

In addition to trees, the `nexml_write()` method also allows to specify
character data as another parameter. The following example uses data
from the comparative phylogenetics R package `geiger` [@Pennell2014].

@@

Note that the NeXML format is well-suited for incomplete data: for instance,
here it does not assume the character matrix has data for every tip in the tree.

@@

The `RNeXML` interface described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies or ontologies. Giving
metadata properties (and where applicable, values) as vocabulary or
ontology terms rather than simple text strings is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics [@Parr2011]. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` interface shields her otherwise, in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

If the prefix used by the metadata property is not among the built-in
ones (which can be obtained using `get_namespaces()`), it has to be
provided along with its URI as the `namespaces` parameter. For example,
the following uses the "[Simple Knowledge Organization System](http://www.w3.org/TR/skos-reference/)" (SKOS) 
vocabulary to add a note to the trees in the `nexml` object:

@@

Because NeXML expresses all metadata using the RDF standard, and stores
them compliant with RDFa, they can be extracted as an RDF graph, queried,
analyzed, and mashed up with other RDF data, local or on the web, using
a wealth of off-the-shelf tools for working with RDF (see @W3C_2014
or @Hartig_2012). Examples for these possibilities are included in the
`RNeXML` SPARQL vignette (a recursive acronym for SPARQL Protocol and
RDF Query Language, see http://www.w3.org/TR/rdf-sparql-query/), and
the package also comes with a demonstration that can be run from R using
the following command: `demo("sparql", "RNeXML")`).

@@

```{r eval = FALSE}
doi <- nexml_publish(birds, repository="figshare")
```

This method reserves a permanent identifier (DOI) on the figshare
repository that can later be made public through the figshare web
interface.  This also acts as a secure backup of the data to a repository
and a way to share with collaborators prior to public release.

@@

# References
